SDSS-IV MaNGA: The Radial Profile of Enhanced Star Formation in Close Galaxy Pairs

### 2021-01-21 maintex.tex
final edits

Among the 6142 MaNGA IFUs considered here, 404 have multiple galaxy objects within the IFU. We further restrict the sample by setting a relative velocity cut of $\Delta v$ $<$ 500 km s$^{-1}$ to remove projected companions. Given the redshift range of the MaNGA sample and the size of MaNGA's IFUs, the maximum projected separation for a companion galaxy in the IFU is $\sim$40 kpc. Again, the galaxy sample is also restricted to galaxies with a stellar mass range of \logm\ $=$ 9.0$-$11.5 and galaxies in the Primary or Secondary MaNGA subsamples. We also require that the galaxies are classified as star forming as described in \S~\ref{sec:data}. These requirements reduce our sample down to 87 star forming MaNGA targets with inside-IFU companions. While the companions to the inside-IFU MaNGA targets contained are within the MaNGA IFUs, we only use the MaNGA targets when building radial profiles. We leave out the identified companions from the analysis because the MaNGA targets are covered by the NSA catalog with geometries and total stellar masses while the companions are not covered by the NSA catalog. \textbf{While we use the identified companions to select galaxy pairs, we only build radial profiles for the MaNGA target galaxies. We do this because the identified companions are not included in the NSA catalog which we use to build the radial profiles for the galaxies.}

Several of the target galaxies in our sample have multiple companions. From the whole set of 404 IFUs, there are 327 pairs, 67 triplets, 7 quadruplets, and 1 quintuplet. In cases where a MaNGA target has multiple viable companions galaxies, we chose the closest companion when measuring the pair's mass ratio and projected separation.

@@

From the NSA catalog's 641,409 galaxies, we find 492 galaxies which are paired to MaNGA targets. After restricting the sample to SFGs, we have another 133 MaNGA targets with paired galaxies outside of the IFU. MaNGA targets which have both an inside-IFU and an outside-IFU pair are left to the inside-IFU sample. This gives us a total of \textbf{169} close pairs, shown in Table \ref{tab:coords}.

@@

To show how the population of galaxy pairs differs from isolated galaxies, we will compare our pair sample to a sample of \textbf{control} galaxies in the MaNGA survey. We build this control sample from the MaNGA target galaxies which have no spectroscopic companions within $r_{\rm p}$ $<$ 50 kpc and $\Delta v$ $<$ 500 km s$^{-1}$ either inside or outside of the IFU. This gives us a control sample of 1888 star forming control galaxies.

It has been shown that the SFR in galaxies is dependent on both the stellar mass and the redshift of the galaxies \citep[e.g.,][]{Noeske:2007}. We will compare our interacting galaxies with \textbf{control} galaxies of a similar stellar mass and redshift, to account for these other dependencies. To do this, we will use two different methods of pair - control comparison.

@@

In the second method, we select a ``tailored'' control sample for each paired galaxy from the full sample of isolated galaxies. We select a subsample of 20 control galaxies for each paired galaxy. We first select all of the control galaxies which are within 0.1 dex in stellar mass limit, \textbf{0.025 in redshift, 20\% \reff}, and within the same MaNGA subsample (e.g. Primary or Secondary) as the paired galaxy. With these requirements, most paired galaxies will find between 20$-$100 control galaxies. Since we want each paired galaxy to be treated in a similar manner, we randomly down-select the total number of acquired control galaxies to 20. A given control galaxy may be selected for multiple pairs by the pipeline; most of the control galaxies are used at least once and the average number of times a given control is reused is between 4$-$6 times. In the cases where a paired galaxy does not initially acquire 20 control galaxies we iteratively expand the stellar mass limit by 0.1 dex, the redshift by 0.025, and the \reff\ by 5\% until at least 20 control galaxies are found. \textbf{39 pairs required 1 extra iteration, 15 pairs required 2 extra iterations, 5 pairs required 3 extra iterations, and 2 pairs required 4 extra iterations. }

@@

We measure the SFR from the extinction corrected H$\alpha$ luminosity, $L_{{\rm H}\alpha}$. We use the SFR formula, Equation \ref{eq:sfr}, from \citet{Murphy:2011} which uses a Kroupa IMF, Solar metallicity, a constant SFR at an age of 100 Myr, and Case-B recombination: 
\begin{equation}\label{eq:sfr}
\frac{\rm{SFR}}{M_{\odot} \, \rm{yr^{-1} kpc^{-2}}} = \frac{L_{\rm H\alpha}}{1.86 \times 10^{41}\, \rm{erg\,s}^{-1}}.
\end{equation}

@@

\textbf{
We check our measurement of the specific star formation rate with the equivalent width of the H$\alpha$ line, \ewha, since it is a known proxy for the sSFR. This is useful as the sSFR is dependent on {\sc SPFIT}'s measurement of the stellar mass while \ewha\ is an observable. As we will show in Section \ref{sec:results}, the \ewha\ is consistent with our measurement of the sSFR.
}

@@

\textbf{We not only use the mass ratio to study how a major merger behaves differently from a minor merger but also to study how the more massive galaxy of a pair behaves differently from the less massive galaxy of the pair. We expect that either component of a similar mass pair will behave in a similar fashion; however, in a minor merger the large central galaxy may not respond to the merger event in the same way as its less massive companion. The two different scenarios can be differentiated with our definition of the mass ratio in equation \ref{eq:dm}; when the MaNGA target is the more massive companion of the pair, its mass ratio is positive and when MaNGA target is the less massive companion of pair the mass ratio is negative. To distinguish between these scenarios, we refer to the more massive galaxy of a pair as the primary companion and the less massive galaxy of a pair the secondary companion.}

@@

\textbf{We also see that the $\Delta$log(sSFR) in the outskirts ($R$ $>$ 1.5 \reff) of the pairs have a dependence on the mass ratio in the bottom right of Figure \ref{fig:ssfr_dm}. Primary companions show a zero to positive enhancement of 0.0$-$0.4 dex to the sSFR in their outskirts, while secondary companions feature a suppression to the sSFR in their outskirts of about 0.0$-$0.4 dex. Again we note that only 37\% of the MaNGA sample is designed to extend beyond 1.5 \reff; however, a similar result has been observed in the hydrodynamical simulations of \citet{Moreno:2015}. }

@@

@@

In Figure \ref{fig:prof_comp} we compare the $\Delta$log(sSFR) profile \textbf{and $\Delta$log(SFR)} as a function of galactocentric radius between our work and previous works. \textbf{\citet{Thorp:2019} studies a sample of 36 post-mergers from the MaNGA sample. The centers of the post-merger galaxies feature a greater level of sSFR enhancement in their centers compared to our galaxies pairs of $\Delta$log(SFR) $=$ 0.40$-$0.45, roughly 0.2 dex higher than our paired galaxies. Between 0.4$-$1.0\reff\ the post-merger galaxies and our pairs are in closer agreement, being 0.05$-$0.1 dex higher than our paired galaxies. Beyond 1.0 \reff\ the post-merger galaxies have a higher $\Delta$log(SFR) by about 0.1$-$0.2 dex.} The heightened level of sSFR enhancement in the centers of post-merger galaxies in comparison to merging galaxies seems consistent with the idea that a second and greater burst of star formation occurs as the two merging galaxies coalesce into a single galaxy.

@@

\section{Physical Scenario}
% General Merger-SFR enhancement
\textbf{
Our results are consistent with the idea that galaxy merger events trigger a burst of star formation in the centers of the paired galaxies. When two galaxy pairs undergo their first pericenter, the tidal torques exerted on the disks of interacting galaxies form barred structures. These barred structures funnel gases into the centers of the galaxies which triggers a burst of new star formation. Eventually, as the pairs separate from each other after the first pericenter, the burst of star formation begins to subside. This is shown in our work and many previous works in Figure \ref{fig:nuc_sep} where the sSFR enhancement is greatest for close separations and falls with wider separations. 
}

% Mass ratio stuff
\textbf{
In agreement with previous works, we find that the strength of the central burst of merger induced star formation is dependent on the relative masses of the two galaxies. Equal mass galaxies see the strongest bursts of star formation while wide mass ratios see weaker bursts of star formation. We further find that the strength of the central burst of star formation differs between the primary and secondary companion of a pair where the less massive secondary companion features a higher level of sSFR enhancement in comparison to its higher mass primary companion. 
}

\textbf{
We also find that there is a difference in the $\Delta$log(sSFR) offsets at wide radii between primary and secondary companions. The primary companion features an enhancement to their sSFR at wide radii while the secondary companion features a suppression to their sSFR at wide radii. The reason for this is unclear; however, this means that the primary companions will experience mass growth across their disks while the secondary companions will see only see substantial mass growth in their centers which will result in more bulge dominated galaxies.
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary and Conclusion}\label{sec:sum}

@@

We thank Cheng Li for our useful discussion on how the sSFR enhancement in galaxy pairs relates to their stellar mass and mass ratio. J.S. and H.F. acknowledge support from the National Science Foundation (NSF) grant AST-1614326. Funding for the Sloan Digital Sky Survey IV has been provided by the Alfred P. Sloan Foundation, the U.S. Department of Energy Office of Science, and the Participating Institutions. SDSS acknowledges support and resources from the Center for High-Performance Computing at the University of Utah. The SDSS web site is www.sdss.org.

>>>>

Among the 6142 MaNGA IFUs considered here, 404 have multiple galaxy objects within the IFU. We further restrict the sample by setting a relative velocity cut of $\Delta v$ $<$ 500 km s$^{-1}$ to remove projected companions. Given the redshift range of the MaNGA sample and the size of MaNGA's IFUs, the maximum projected separation for a companion galaxy in the IFU is $\sim$40 kpc. Again, the galaxy sample is also restricted to galaxies with a stellar mass range of \logm\ $=$ 9.0$-$11.5 and galaxies in the Primary or Secondary MaNGA subsamples. We also require that the galaxies are classified as star forming as described in \S~\ref{sec:data}. These requirements reduce our sample down to 54 star forming MaNGA targets with inside-IFU companions. \textbf{While we use the identified companions to select galaxy pairs, we only build radial profiles for the MaNGA target galaxies. We do this because the identified companions are not included in the NSA catalog which we use to build the radial profiles for the galaxies.}

Several of the target galaxies in our sample have multiple companions. From the whole set of 404 IFUs, there are 327 pairs, 67 triplets, 7 quadruplets, and 1 quintuplet. In cases where a MaNGA target has multiple viable companions galaxies, we chose the closest companion when measuring the pair's mass ratio and projected separation.  \textbf{We show the whole 404 inside-IFU pair sample in Table \ref{tab:coords}.}

@@

From the NSA catalog's 641,409 galaxies, we find 492 galaxies which are paired to MaNGA targets. After restricting the sample to SFGs, we have another 115 MaNGA targets with paired galaxies outside of the IFU. MaNGA targets which have both an inside-IFU and an outside-IFU pair are left to the inside-IFU sample.

@@

To show how the population of galaxy pairs differs from isolated galaxies, we will compare our pair sample to a sample of \textbf{control} galaxies in the MaNGA survey. We build this control sample from the MaNGA target galaxies which have no spectroscopic companions within $r_{\rm p}$ $<$ 50 kpc and $\Delta v$ $<$ 500 km s$^{-1}$ either inside or outside of the IFU. This gives us a control sample of 1830 star forming control galaxies.

It has been shown that the SFR in galaxies is dependent on both the stellar mass and the redshift of the galaxies \citep[e.g.,][]{Noeske:2007}. We will compare our interacting galaxies with \textbf{control} galaxies of a similar stellar mass, redshift, and radial size, to account for these other dependencies. To do this, we will use two different methods of pair - control comparison.

@@

In the second method, we select a ``tailored'' control sample for each paired galaxy from the full sample of isolated galaxies. We select a subsample of 20 control galaxies for each paired galaxy. We first select all of the control galaxies which are within 0.1 dex in stellar mass, \textbf{0.025 in redshift, 20\% in effective radii}, and within the same MaNGA subsample (e.g. Primary or Secondary) as the paired galaxy. With these requirements, most paired galaxies will find between 20$-$60 control galaxies. Since we want each paired galaxy to be treated in a similar manner, we randomly down-select the total number of acquired control galaxies to 20. A given control galaxy may be selected for multiple pairs by the pipeline; most of the control galaxies are used at least once and the average number of times a given control is reused is between 2$-$4 times. In the cases where a paired galaxy does not initially acquire 20 control galaxies we iteratively expand the stellar mass limit by 0.1 dex, the redshift by 0.025, and the \reff\ by 5\% until at least 20 control galaxies are found. \textbf{39 pairs required 1 extra iteration, 15 pairs required 2 extra iterations, 5 pairs required 3 extra iterations, and 2 pairs required 4 extra iterations. }

@@

We measure the SFR from the extinction corrected H$\alpha$ luminosity, $L_{{\rm H}\alpha}$. We use the SFR formula, Equation \ref{eq:sfr}, from \citet{Murphy:2011} which uses a Kroupa IMF, Solar metallicity, a constant SFR at an age of 100 Myr, and Case-B recombination: 
\begin{equation}\label{eq:sfr}
\frac{\rm{SFR}}{M_{\odot} \, \rm{yr^{-1}}} = \frac{L_{\rm H\alpha}}{1.86 \times 10^{41}\, \rm{erg\,s}^{-1}}.
\end{equation}

@@

\textbf{
We check our measurement of the specific star formation rate with the equivalent width of the H$\alpha$ line, \ewha, since it is a known proxy for the sSFR. This is useful as the sSFR is dependent on {\sc SPFIT}'s measurement of the stellar mass while \ewha\ is an observable. As we will show in Section \ref{sec:results}, the radial profiles of \ewha\ is consistent with radial profiles of the sSFR.
}

@@

\textbf{We not only use the mass ratio to study how a major merger behaves differently from a minor merger but also to study how the more massive galaxy of a pair behaves differently from the less massive galaxy of the pair. We expect that either component of a similar mass pair will behave in a similar fashion; however, in a minor merger the large central galaxy may not respond to the merger event in the same way as its less massive companion. The two different scenarios can be differentiated with our definition of the mass ratio in equation \ref{eq:dm}; when the MaNGA target is the more massive companion of the pair, its mass ratio is positive and when MaNGA target is the less massive companion of pair the mass ratio is negative. To distinguish between these scenarios, we refer to the more massive galaxy of a pair as the primary companion and the less massive galaxy of a pair as the secondary companion.}

@@

\textbf{We also see that the $\Delta$log(sSFR) in the outskirts ($R$ $>$ 1.5 \reff) of the pairs have a dependence on the mass ratio in the bottom right of Figure \ref{fig:ssfr_dm}. Primary companions show a zero to positive enhancement of 0.0$-$0.4 dex to the sSFR in their outskirts, while secondary companions feature a suppression to the sSFR in their outskirts of about 0.0$-$0.4 dex. Again we note that only 37\% of the MaNGA sample is designed to extend beyond 1.5 \reff; however, a similar result has been observed in the hydrodynamical simulations of \citet{Moreno:2015} and \citet{Moreno:2020}. }

@@

\section{Physical Scenario}
% General Merger-SFR enhancement
\textbf{
Our results are consistent with the idea that galaxy merger events trigger a burst of star formation in the centers of the paired galaxies. In this model, when two galaxy pairs undergo their first pericenter, the tidal torques exerted on the disks of interacting galaxies form barred structures. These barred structures funnel gases into the centers of the galaxies which triggers a burst of new star formation. Eventually, as the pairs separate from each other after the first pericenter, the burst of star formation begins to subside. This is shown in our work and many previous works in Figure \ref{fig:nuc_sep} where the sSFR enhancement is greatest for close separations and falls with wider separations. 
}

% Mass ratio stuff
\textbf{
We find that the strength of the central burst of merger induced star formation is dependent on the relative masses between the two galaxies. Equal mass galaxies see the strongest bursts of star formation while wide mass ratios see weaker bursts of star formation. We further find that the strength of the central burst of star formation differs between the primary and secondary companion of a pair where the less massive secondary companion features a higher level of sSFR enhancement in comparison to its higher mass primary companion. 
}

\textbf{
\citet{Moreno:2020} uses galaxy merger simulations to study the origin of the enhanced star formation in galaxy pairs, whether its an increase in the star formation efficiency, defined as the ratio between the SFR and the mass of cold-dense gas, or its an increase in the availability of cold dense gas. They find that the star formation in secondary galaxies is evenly split between being efficiency driven or fuel driven systems while primary galaxies are more likely to be fuel driven systems (71\%). \citet{Moreno:2020} also finds that that secondary galaxies feature higher levels of star formation enhancement in comparison to primary galaxies, which is in agreement with what we find in this work. This indicates that the reason why primary galaxies behave differently from secondary galaxies may be due to a difference in the physical mechanism which drives the enhanced star formation. 
}

\textbf{
We also find that there is a difference in the $\Delta$log(sSFR) offsets at wide radii between primary and secondary companions. The primary companion features an enhancement to their sSFR at wide radii while the secondary companion features a suppression to their sSFR at wide radii. The reason for this is unclear; however, this means that the primary companions will experience enhanced stellar mass growth across their disks while the secondary companions will see only see substantial stellar mass growth in their centers which will result in more bulge dominated galaxies.
}

@@

In Figure \ref{fig:prof_comp} we compare the $\Delta$log(sSFR) profile and $\Delta$log(SFR) as a function of galactocentric radius between our work and previous works. \textbf{\citet{Thorp:2019} studies a sample of 36 post-mergers from the MaNGA sample. The centers of the post-merger galaxies feature a greater level of sSFR enhancement in their centers compared to our galaxies pairs of $\Delta$log(SFR) $=$ 0.40$-$0.45, roughly 0.2 dex higher than our paired galaxies. Between 0.4$-$1.0\reff\ the post-merger galaxies and our pairs are in closer agreement, being 0.05$-$0.1 dex higher than our paired galaxies. Beyond 1.0 \reff\ the post-merger galaxies have a higher $\Delta$log(SFR) by about 0.1$-$0.2 dex.} The heightened level of sSFR enhancement in the centers of post-merger galaxies in comparison to merging galaxies seems consistent with the idea that a second and greater burst of star formation occurs as the two merging galaxies coalesce into a single galaxy.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary and Conclusion}\label{sec:sum}

@@

\textbf{We thank the anonymous referee for the useful comments and suggestions.} We also thank Cheng Li for our useful discussion on how the sSFR enhancement in galaxy pairs relates to their stellar mass and mass ratio. J.S. and H.F. acknowledge support from the National Science Foundation (NSF) grant AST-1614326. Funding for the Sloan Digital Sky Survey IV has been provided by the Alfred P. Sloan Foundation, the U.S. Department of Energy Office of Science, and the Participating Institutions. SDSS acknowledges support and resources from the Center for High-Performance Computing at the University of Utah. The SDSS web site is www.sdss.org.
